### 2014-07-26 Algorithms/MST/mst.tex
Finished up Kruskal's and Prim's section

\begin{lstlisting}
edges = [(C, D, 1), (C, E, 1), (D, E, 2), (A, B, 3), (B, F, 4), (E, F, 4), 
         (B, C, 5), (C, F, 5), (A, F, 6)]
\end{lstlisting}

Now we begin iterating through the edges to build the spanning tree.
The first edge in the list is \li{(C,D,1)}.
The root for \li{C} is \li{C} and the root for \li{D} is \li{D}, so we add this edge to the tree.
The tree becomes
\begin{lstlisting}
[('C', 'D', 1)]
\end{lstlisting}
We then change the root node of \li{D} to be \li{C}.
The dictionary of root nodes now looks like

@@

Now we process the next edge, \li{(C, E, 1)}.
The root node of \li{C} is \li{C}, and the root node of \li{E} is \li{E}, so adding this edge does not create a cycle.
We add this edge into the tree, so the tree is now

\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1)]
\end{lstlisting}
Then we change the root of \li{E} so that it is \li{C}. The dictionary is now

@@

The next edge is \li{(D, E, 2)}.
The root node of the tree containing \li{D} is \li{C} and the root node of the tree containing \li{E} is \li{C}, so these nodes are already connected, so we do not add this edge to the spanning tree.

The next edge is \li{(A, B, 3)}.
The root node for \li{A} is \li{A}, and the root node for \li{B} is \li{B}, so we add the edge to the tree and update the dictionary.
The dictionary becomes

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'F'}
\end{lstlisting}
The tree becomes
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3)]
\end{lstlisting}

The next edge is \li{(B, F, 4)}.
The root node for \li{B} is \li{A} and the root node for \li{F} is \li{F}, so we add the edge to the tree and change the root node of \li{F} to be \li{A}.

The dictionary becomes

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'A'}
\end{lstlisting}
The tree becomes
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4)]
\end{lstlisting}
The next edge is \li{(E, F, 4)}.
The root node for \li{E} is \li{C} and the root node for \li{F} is \li{A}.
We add the edge to the tree and end the algorithm since the spanning tree now has five edges.

The final spanning tree looks like
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4), ('E', 'F', 4)]
\end{lstlisting}
If we were to continue the algorithm, we would update the root node of \li{A} to be \li{C}. The dictionary would then look like

@@

One of the keys to this algorithm is to make sure we are not adding cycles into the spanning tree. A way to think about how the algorithm does this is by picturing it building smaller trees that have a common root node, and then as the algorithm progresses, all of these smaller trees join to have the same root node. To find the root node of the tree, we look through our dictionary until we find a node that points to itself.
For example, to find the root node of \li{F} we first need to get the value for \li{F} from the dictionary.  This is \li{A}.
Then we get the value for \li{A} from the dictionary which is \li{C}.
Since the value of \li{C} in the dictionary is itself, then \li{C} is the root node of the graph containing \li{F}.
It is necessary to trace through the dictionary in this manner each time because we avoid iterating over all of the nodes and updating their root values everytime.

@@

% Note: The comments in the solutions match the pseudocode.
% When making changes, please keep that in mind.

@@

\begin{problem}
Implement Kruskal's algorithm.
Test your algorithm on random symmetric arrays.
You can generate a random symmetric array by multiplying a random array by its transpose.
Use the data from MSTdata.npy to test your tree.
Use np.load("MSTdata.npy") to get it.
Use the \li{formChanger} function below to put it in the right form.
\begin{lstlisting}
def formChanger(oldData):
    newData=[]
    for i in oldData: newData.append((i[0],i[1],int(i[2])))
    return newData
\end{lstlisting}
\end{problem}
\section*{Prim's algorithm}

@@

Again, consider the example shown in Figure 7.4.
We first initialize a dictionary with all the nodes as keys in order to track which nodes have not been processed.
At the beginning it will be \li{\{A:False, B:False, C:False, D:False, E:False, F:False\}}.
We then form the dictionary that maps each node to the edges that contain it.
Since we will already know one node of each edge while we use the dictionary, we only need to store the node that is not being looked up.
It should end up looking like \li{\{A:[(B, 3), (F, 6)], B:[(A, 3), (C, 5), (F, 4)], C:[(B, 5), (D, 1), (E, 1), (F, 5)], D:[(C, 1), (E, 2)], E:[(C, 1), (D, 2), (F, 4)], F:[(A, 6), (B, 4), (C, 5), (E, 4)]\}}.
We will also initialize an empty dictionary to track the shortest edges that run between nodes we have processed and nodes that we haven't.
As we iterated over the edges in our initialization step, we are also able to find the shortest edge.
In this case, it is \li{(D, C, 1)}.
Let's start with \li{(D, C, 1)} and initialize our tree as the list \li{[(D, C, 1)]}.
Next, we mark \li{D} and \li{C} as processed in the dictionary that tracks which nodes have been processed.
We now start to build our dictionary of nodes that are one edge away from our processed nodes.
In this case, after adding the shortest edges between processed and unprocessed nodes to the dictionary, the dictionary becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(C, F, 5)\}}.
Notice we did not include \li{E:(D, E, 2)} because there is a shorter edge to \li{E} from \li{C}.

Of the edges in the dictionary of edges that can be processed next, the shortest is \li{(C, E, 1)}, so we add that edge to the tree and mark \li{E} as processed.
With \li{E} being processed, we can now reach \li{F} at a cost of 4.
After making this change, the dictionary of edges to process becomes \li{\{B:(C, B, 5), E:(C, E, 1), F:(E, F, 4)\}}.
Since \li{E} no longer needs to be processed, we can remove it from consideration.
So this dictionary now becomes \li{\{B:(C, B, 5), F:(E, F, 4)\}}.

Of the edges to be processed next, \li{(E, F, 4)} is the shortest, so we add it to the tree and mark \li{F} as processed.
After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{B:(F, B, 4), A:(F, A, 6)\}}.

Of the edges to be processed next, \li{(F, B, 4)} is the shortest, so we add it to the tree and mark \li{B} as processed.
After performing the appropriate modifications to the dictionary of edges to be processed next, it becomes \li{\{A:(B, A, 3)\}}.

The only edge to be considered is \li{(B, A, 3)}, so we add it to the tree.
The tree is long enough that it it spans the nodes, so the algorithm is finished.

Here's pseudocode for a version of Prim's algorithm.
While it is not a perfectly optimized version, it is pretty good.

@@

\item Initialize a dictionary to track which nodes have been processed.

\item Initialize an empty dictionary of lists to track the edges containing each node.

\item Fill the edge list.
	Be sure to add each edge to the list corresponding to both of its nodes.

\item Get the first edge to add (The shortest edge from any given node is a good pick).

\item Mark the nodes in the first edge as processed.

\item Initialize the tree to be the list containing the first edge.

@@

	\item If that node isn't in the dictionary, set its value to be the edge passed to the functions.

	\item If it is in the dictionary already, set its value to be the shorter of the edges being processed and the edges already in the dictionary.

@@

Many graph packages have functions for finding the minimal spanning tree. One of these is NetworkX. You begin as follows
\begin{lstlisting}
import networkx as nx
G=nx.Graph()
\end{lstlisting}
You can add a node $x$ with \li{G.add_node(x)} and add a edge from $x$ to $y$ with weight n with \li{G.add_edge(x,y,weight=n)}. Then \li{nx.minimum_spanning_tree(G)} returns the minimal spanning tree using Kruskals algorithm.

\begin{problem}
Use NetworkX to find the minimal spanning tree of the data in MSTdata.npy.
Compare the output with that of your algorithm.
Create a function which prints your comparison.
\end{problem}

@@

One application of Minimal Spanning Trees (MSTs) is image segmentation.
Kruskal's algorithm is especially good at this.
You can convert an image into graph. Each pixel is a vertex and you define weights between the nodes. You then build the minimal spanning tree. If you take away the edge with the greatest weight you will get two forests (removing an edge from any graph with no cycles will create two forests). Then you can continue doing this to each forest to split up those forests. Each forest corresponds to a segment of the image.
Let $k$ be the number of divisions that is wanted and $n$ be the number of nodes.
Kruskal's algorithm is performed until $n-(k+1)$ edges are added which is the same as taking out the $k$ edges with greatest weights.

@@

The provided Kruskal's algorithm takes as inputs the list of nodes, the list of edges and the number of divisions desired.
The number of divisions often has to be higher than the actual number that is needed because sometimes one or two pixels form a division because the difference between them and the pixels around them is so great.
You will have to adjust the number of divisions until the desired result is found.  See Figure 9.1.

@@

\vfill
\begin{figure}[ht]
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg1.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg2.jpg}
\end{minipage}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg3.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg4.jpg}
\end{minipage}
\caption{The original image is in the top left hand corner. The three larges segments are shown in the other corners. The original image was 498x498 and 50000 divisions were used.}
\end{figure}
\vfill

>>>>

\begin{lstlisting}
edges = [('C', 'D', 1), ('C', 'E', 1), ('D', 'E', 2), ('A', 'B', 3), 
	     ('B', 'F', 4), ('E', 'F', 4), ('B', 'C', 5), ('C', 'F', 5), 
	     ('A', 'F', 6)]
\end{lstlisting}

Now we begin iterating through the edges to build the spanning tree.
The first edge in the list is \li{('C', 'D',1)}.
The root for \li{'C'} is \li{'C'} and the root for \li{'D'} is \li{'D'}, so we add this edge to the tree.
The tree becomes:
\begin{lstlisting}
[('C', 'D', 1)]
\end{lstlisting}
We then change the root node of \li{'D'} to be \li{'C'}.
The dictionary of root nodes now looks like:

@@

Now we process the next edge, \li{('C', 'E', 1)}.
The root node of \li{'C'} is \li{'C'}, and the root node of \li{'E'} is \li{'E'}, so adding this edge does not create a cycle.
We add this edge into the tree, so the tree is now:

\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1)]
\end{lstlisting}
Then we change the root of \li{'E'} so that it is \li{'C'}. The dictionary is now:

@@

The next edge is \li{('D', 'E', 2)}.
The root node of the tree containing \li{'D'} is \li{'C'} and the root node of the tree containing \li{'E'} is \li{'C'}, so these nodes are already connected, so we do not add this edge to the spanning tree.

The next edge is \li{('A', 'B', 3)}.
The root node for \li{'A'} is \li{'A'}, and the root node for \li{'B'} is \li{'B'}, so we add the edge to the tree and update the dictionary.
The dictionary becomes:

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'F'}
\end{lstlisting}
The tree becomes:
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3)]
\end{lstlisting}

The next edge is \li{('B', 'F', 4)}.
The root node for \li{'B'} is \li{'A'} and the root node for \li{'F'} is \li{'F'}, so we add the edge to the tree and change the root node of \li{'F'} to be \li{'A'}.

The dictionary becomes:

\begin{lstlisting}
{'A': 'A', 'B':'A', 'C':'C', 'D':'C', 'E':'C', 'F':'A'}
\end{lstlisting}
The tree becomes:
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4)]
\end{lstlisting}
The next edge is \li{('E', 'F', 4)}.
The root node for \li{'E'} is \li{'C'} and the root node for \li{'F'} is \li{'A'}.
We add the edge to the tree and end the algorithm since the spanning tree now has five edges.

The final spanning tree looks like:
\begin{lstlisting}
[('C', 'D', 1), ('C', 'E', 1), ('A', 'B', 3), ('B', 'F', 4), ('E', 'F', 4)]
\end{lstlisting}
If we were to continue the algorithm, we would update the root node of \li{'A'} to be \li{'C'}. The dictionary would then look like:

@@

One of the keys to this algorithm is to make sure we are not adding cycles into the spanning tree. A way to think about how the algorithm does this is by picturing it building smaller trees that have a common root node, and then as the algorithm progresses, all of these smaller trees join to have the same root node. To find the root node of the tree, we look through our dictionary until we find a node that points to itself.
For example, to find the root node of \li{'F'} we first need to get the value for \li{'F'} from the dictionary.  This is \li{'A'}.
Then we get the value for \li{'A'} from the dictionary which is \li{'C'}.
Since the value of \li{'C'} in the dictionary is itself, then \li{'C'} is the root node of the graph containing \li{'F'}.
It is necessary to trace through the dictionary in this manner each time because we avoid iterating over all of the nodes and updating their root values everytime.

@@

% Note: The comments in the solutions match the outline.
% When making changes, please keep that in mind.

@@

\begin{problem}
Implement Kruskal's algorithm.
Use the data from MSTdata.npy to test your tree. To load the data use
\begin{lstlisting}
np.load("MSTdata.npy")
\end{lstlisting}
Use the \li{formChanger} function below to put it in the right form.
\begin{lstlisting}
def formChanger(oldData):
    newData=[]
    for i in oldData: newData.append((i[0],i[1],int(i[2])))
    return newData
\end{lstlisting}
\end{problem}
\section*{Prim's algorithm}

@@

Again, consider the example shown in Figure \ref{mst:graph4} .
We first initialize a dictionary with all of the nodes as keys in order to track which nodes have not been processed.
At the beginning it will look like:
\begin{lstlisting}
{'A': False, 'B': False, 'C': False, 'D': False, 'E': False, 'F': False}
\end{lstlisting}
We set all of the values to \li{'False'} since none of the nodes have been processed yet.

We then form another dictionary that maps each node to the edges that contain it. For example, the node \li{'A'} is contained in the edge \li{('A', 'B', 3)} and the edge \li{('A', 'F', 6)}, so we map \li{'A'} to these edges. Since we already know that \li{'A'} is in these edges, we only need to store the other nodes and weights, which in this case would be \li{('B', 3)} and \li{('F', 6)}.
The dictionary should end up looking like:
\begin{lstlisting}
{'A': [('B', 3), ('F', 6)],
 'B': [('A', 3), ('C', 5), ('F', 4)],
 'C': [('B', 5), ('D', 1), ('E', 1), ('F', 5)],
 'D': [('C', 1), ('E', 2)],
 'E': [('C', 1), ('D', 2), ('F', 4)],
 'F': [('A', 6), ('B', 4), ('C', 5), ('E', 4)]}
\end{lstlisting}
We also initialize an empty dictionary to track the shortest edges that run between nodes we have processed and nodes we haven't. In total, we should now have three different dictionaries.

Next we find the first edge to add to our minimum spanning tree. We can choose this to be the shortest edge, or edge with the smallest weight, from any given node. In this case, we choose \li{('D', 'C', 1)} and initialize our spanning tree as the list \li{[('D', 'C', 1)]}.
We then mark \li{'D'} and \li{'C'} as processed in the dictionary that tracks which nodes have been processed. We do this by changing their values from \li{'False'} to \li{'True'}.

Next we start to build our dictionary of nodes that are one edge away from our processed nodes. This dictionary is the one we initialized above as an empty dictionary. In this case, the nodes that are one edge away from \li{'D'} and \li{'C'} are \li{'B'}, \li{'F'}, and \li{'E'}. Notice that \li{'E'} shares an edge with both \li{'D'} and \li{'C'}. To deal with this, we choose the edge connected to \li{'E'} with the smallest weight, which would be the edge corresponding to \li{'C'}. 
Our dictionary then becomes:
\begin{lstlisting}
{'B':('C', 'B', 5), 'E':('C', 'E', 1), 'F':('C', 'F', 5)}
\end{lstlisting}
Again, notice that we did not include the key/value pair \li{'E':('D', 'E', 2)} because there is a shorter edge to \li{'E'} from \li{'C'}.

Of the edges in this dictionary that can be processed next, the shortest is \li{('C', 'E', 1)}, so we add that edge to the tree and mark \li{'E'} as processed.
With \li{'E'} being processed, we can now reach \li{'F'} at a smaller cost; four now, instead of five.
After making these changes, the dictionary of edges to process becomes:

\begin{lstlisting}
{'B':('C', 'B', 5), 'E':('C', 'E', 1), 'F':('E', 'F', 4)}
\end{lstlisting}

Since \li{'E'} no longer needs to be processed, we can remove it from consideration.
So this dictionary now becomes:

\begin{lstlisting}
{'B':('C', 'B', 5), 'F':('E', 'F', 4)}
\end{lstlisting}

Of the edges to be processed next, \li{('E', 'F', 4)} is the shortest, so we add it to the tree and mark \li{'F'} as processed.
After performing the appropriate modifications to the dictionary of edges to be processed next, we have:
\begin{lstlisting}
{'B':('F', 'B', 4), 'A':('F', 'A', 6)}
\end{lstlisting}

Of the edges to be processed next, \li{('F', 'B', 4)} is the shortest, so we add it to the tree and mark \li{'B'} as processed.
We now have:
\begin{lstlisting}
{'A':('B', 'A', 3)}
\end{lstlisting}

The only edge to be considered is \li{('B', 'A', 3)}, so we add it to the tree.
The tree is long enough that it spans the nodes, so the algorithm is finished.

Here's an outline for a version of Prim's algorithm.
While it is not a perfectly optimized version, it is still effective.

@@

\item Initialize a dictionary to track which nodes have been processed. This will look like the dictionary shown above with nodes for keys and values set to \li{'False'}.

\item Create a dictionary that has for its key/value pairs all of the nodes for keys and the lists of edges each node is a part of for values. To do this first:
	\begin{itemize}
	\item Initialize an empty dictionary of lists.

	\item Iterate through the edges of the graph to fill the lists in the above dictionary.
		Be sure to add each edge to the lists corresponding to both of its nodes.
	\end{itemize}
\item Get the first edge to add. The shortest edge from any given node is a good pick.
\item Mark the nodes in the first edge as processed.

\item Initialize the tree to be a list containing the first edge.

@@

	\item If that node isn't in the dictionary, add it, and set its value to be the edge passed to the function.

	\item If it is in the dictionary already, set its value to be the shorter of the two edges -  either the edge being processed currently or the edge that is already in the dictionary.

@@

Many graph packages have functions for finding the minimum spanning tree. One of these is NetworkX. You begin as follows:
\begin{lstlisting}
import networkx as nx
G=nx.Graph()
\end{lstlisting}
You can add a node $x$ with \li{G.add_node(x)} and add a edge from $x$ to $y$ with weight n with \li{G.add_edge(x,y,weight=n)}. Then \li{nx.minimum_spanning_tree(G)} returns the minimum spanning tree using Kruskal's algorithm.

\begin{problem}
Use NetworkX to find the minimum spanning tree of the data in MSTdata.npy.
Compare the output with that of your algorithm.
Create a function which prints your comparison.
\end{problem}

@@

One application of MSTs is image segmentation. Image segmentation is the process that takes an image and seperates it into similar parts.
Kruskal's algorithm is especially good at this. To use Kruskal's we first convert the image we want to segment into a graph. We consider each pixel to be a node and define weights between the nodes. You then build the minimum spanning tree. If you take away the edge with the greatest weight you will get two forests (removing an edge from any graph with no cycles will create two forests). Then you can continue doing this to each forest to split up those forests. Each forest corresponds to a segment of the image.
Let $k$ be the number of divisions that is wanted and $n$ be the number of nodes.
Kruskal's algorithm is performed until $n-(k+1)$ edges are added which is the same as taking out the $k$ edges with greatest weights.

@@

The provided Kruskal's algorithm takes as inputs the list of nodes, the list of edges and the number of divisions desired.
The number of divisions often has to be higher than the actual number that is needed because sometimes one or two pixels form a division because the difference between them and the pixels around them is so great.
You will have to adjust the number of divisions until the desired result is found.  See Figure 5.6.

@@

\vfill
\begin{figure}[ht]
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg1.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg2.jpg}
\end{minipage}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg3.jpg}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.47\linewidth}
\centering
\includegraphics[width=\textwidth]{MSTseg4.jpg}
\end{minipage}
\caption{The original image is in the top left hand corner. The three larges segments are shown in the other corners. The original image was 498x498 and 50000 divisions were used.}
\label{mst:graph6}
\end{figure}
\vfill
